Learning how to act: making good decisions with machine learning

### 2017-08-26 thesis/thesis_map.tex
creates extended version of nip paper for arXiv, restructuring work on thesis

\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{What if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs. The former aims to identify the effect of a known cause. The latter can be viewed as identiying causes of an effect. They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}. However, state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined. The goal of identifying the causes of an effect can be formalised within the graphical causal model framework. Solving this problem is certainly much more challanging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables. These practical difficulties may well be overwealming, particularly in fields such as social science and economics where datasets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively. However, this does not mean that the problem of identifying causes of effects is ill-posed in principle. It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich liturature which we review briefly in section \ref{sec:causal-discovery}.

@@

\chapter{Causal inference from observational data}
\label{chap:causality}

@@

Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. To do this we need terminology to describe actions and how we anticipate the system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks. In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.

@@

\section{Causal Bayesian networks}

@@

\subsection{Can't everything be encoded in a Bayesian model?}
Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention. The power of defining an intervention in this way stems from the number of things that are invariant between the two processes. All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.

@@

\subsection{Limitations of causal Bayesian networks}
A number of criticisms have been levelled at this approach to modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}. However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions. For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter. But, what if some people failed to take the prescribed treatment? We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}. Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.

@@

\section{Counterfactuals}
\label{sec:counterfactuals}

@@

\section{Structural Equation models}
\label{sec:SEM}

@@

\section{Comparing and unifying the models}

@@

\section{The Do Calculus} 
\label{sec:do_calculus}

@@

\section{Estimation}
\subsection{Defining causal effects} 
So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data. This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set. From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.

@@

\subsection{Estimating causal effects by adjusting for confounding variables}
\label{sec:causal_effect_under_ignorability}
Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors. Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds. Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds. This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}. The setting is also referred to as unconfounded.

>>>>

\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{What if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs. The former aims to identify the effect of a known cause. The latter can be viewed as identiying causes of an effect. They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}. However, state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined. The goal of identifying the causes of an effect can be formalised within the graphical causal model framework. Solving this problem is certainly much more challanging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables. These practical difficulties may well be overwealming, particularly in fields such as social science and economics where datasets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively. However, this does not mean that the problem of identifying causes of effects is ill-posed in principle. It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich liturature which we review briefly in section \ref{sec:causal-discovery}.
\todo{reference a review paper on causal discovery here}

@@

\chapter{Learning from observational data}
\label{chap:causality}

The goal of causal inference is to learn the effect of taking an action. We can do this directly via experimental approaches, however any given agent only has a limited capacity to manipulate the world. We are generating and storing data on almost every aspect of our lives at an unprecedented rate. As we incorporate sensors and robotics into our cities, homes, cars, everyday products and even our bodies, the breadth and scale of this data will only increase. However, only a tiny fraction of this data will be generated in a controlled way with the specific goal of answering a single question. An agent that can only learn from data when it had explicit control (or perfect knowledge of) the process by which that data was generated will be severly limted. This makes it critical that we develop effective methods that us to predict the outcome of an intervention in some system from observing it. This is the problem of observational causal inference.

@@

As previusly mentioned, observational causal inference is

Observational causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. As previuosly mentioned, this is a form of transfer learning; we need to infer properties of the system post-intervention from observations of the system pre-intervention. Doing this requires some assumptions about how these two systems are related, or other words, a way of describing actions and how we anticipate the system will respond to them. Three key approaches have emerged: counterfactuals, structural equation models and causal Bayesian networks.

counterfactuals

MOTIVATE WHY I AM DESCRIBING THEM ALL SUMMARISE WHAT EACH OFFERS.

In this chapter we will examine the problems these approaches allow us to solve, the assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms needed to tackle more interesting problems later on.

@@

\subsection{Causal Bayesian networks}

@@

\subsubsection{Can't everything be encoded in a Bayesian model?}
Formalising the definition of an intervention within the framework of causal graphical models provides us with an explicit mechanism to map information from one data generating process, the system pre-intervention, to another, the system post-intervention. The power of defining an intervention in this way stems from the number of things that are invariant between the two processes. All the (conditional) distributions for variables in the graph that were not directly set by the intervention are assumed not be changed by it.

@@

\subsubsection{Limitations of causal Bayesian networks}
A number of criticisms have been levelled at this approach to modelling causality. One is that the definition of an intervention only in terms of setting the value of one or more variables is too precise and that any real world intervention will affect many variables in complex and non-deterministic ways \citep{rickles2009causality,cartwright2007hunting}. However, by augmenting the causal graph with additional variables that model how interventions may take effect, the deterministic do operator can model more complex interventions. For example, in the drug treatment case, we assumed that all subjects complied, taking the treatment or placebo as assigned by the experimenter. But, what if some people failed to take the prescribed treatment? We can model this within the framework of deterministic interventions by adding a node representing what they were prescribed (the intervention) which probabilistically influences the treatment they actually receive, see figure \ref{fig:randomized_imperfect_compliance}. Note that the fact that we no longer directly assign the treatment opens the possibility that an unobserved latent variable could affect both the actual treatment taken and the outcome.

@@

\subsection{Counterfactuals}
\label{sec:counterfactuals}

@@

\subsection{Structural Equation models}
\label{sec:SEM}

@@

\subsection{Comparing and unifying the models}

@@

\subsection{The Do Calculus} 
\label{sec:do_calculus}

@@

\subsection{Estimation}
\subsubsection{Defining causal effects} 
So far we have described causal effect estimation in term of identifying the interventional distribution $\P{Y|do(X)}$ from observational data. This interventional distribution is in fact a family of distributions parameterised by the value, $x$, to which the treatment variable $X$ is set. From a decision theoretic viewpoint, we can select an optimal action $x$ by specifying a utility function $\mathcal{U}:y \in \mathcal{Y} \rightarrow \mathbb{R}$ that assigns a value to each outcome $y$ and then selecting the action that maximises the expected utility.

@@

\subsubsection{Estimating causal effects by adjusting for confounding variables}
\label{sec:causal_effect_under_ignorability}
Probably the two most frequently applied approaches to estimating causal effects from observational data are instrumental variables and adjusting for confounding factors. Instrumental variables correspond to the graph in figure \ref{fig:nid2_instrumental_variable}, which is not identifiable without parametric assumptions, however they can provide tight bounds. Adjusting for confounding equates to identifying a set of variables $\boldsymbol{Z}$ such that the ignorability assumption discussed in section \ref{sec:counterfactuals} holds. This corresponds to a simple graphical test known as the backdoor criterion \citep{Pearl2000}. The setting is also referred to as unconfounded.
